---
title: "Nested Sampling with Ernest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested Sampling with Ernest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r srr-tags, eval = FALSE, echo = FALSE}
#' roxygen_block_name
#' 
#' @srrstats {G1.3} Vignette presents consistent statistical terminology to be
#' used throughout ernest.
#' @srrstats {BS1.2, BS1.2b} Contains examples for specifying priors.
#' @srrstats {BS1.3, BS1.3a} Describes how to save and continue previous runs
#' with `generate()`.
```

```{r setup, message = FALSE}
library(ernest)
library(brms)
library(truncnorm)
library(ggplot2)
library(posterior)
```

# Nested Sampling with Ernest

Bayesian inference relies on using probabilities to describe a researcher's degree of belief in a statistical model ($M$) of some observed data ($D$). This relationship is captured through Bayes' theorem [@ashton2022]:

$$
\Pr(\theta|D,M) = \frac{\Pr(D|\theta, M) \Pr(\theta|M)}{\Pr(D|M)} = \frac{L(\theta)\pi(\theta)}{Z}
$$
where

-   $\Pr(\theta|M) = \pi(\theta)$ is the *prior*, representing the researcher's current estimate of $M$'s unknown parameters $\theta$ before considering the observations;

-   $\Pr(D|\theta,M) = L(\theta)$ is the *likelihood*, representing the the probability of generating $D$ given $M$ and some parameters $\theta$;

-   $\Pr(\theta|D,M) = \Pi(\theta)$ is the *posterior*, representing our estimate of $\theta$ after it has been updated with $D$, and;

-   $\Pr(D|M) = Z$ is the *model evidence*, *evidence*, or *marginal likelihood*.

For most data and models, the posterior and evidence is often intractable and must be estimated using Bayesian computation, where we use a model's likelihood and prior to generate estimates of the desired quantity. Historically, these methods focus on estimating $\Pi(\theta)$: for example, the popular Monte Carlo Markov Chain $MCMC$ method purports to construct a series of estimates of $\theta$ that successfully approximates the posterior. Consider this example from the brms [@bürkner2017] package, where we use an MCMC to estimate the parameters of a Poisson regression model:

```{r}
data("epilepsy")
fit1 <- brm(
  count ~ zBase * Trt,
  prior = prior("normal(0, 10)", class = "b") + prior("normal(0, 10)", class = "Intercept"),
  data = epilepsy,
  family = poisson(),
  refresh = 0
)
summary(fit1)
```

Where the posterior is most useful for estimating our model's parameters, model evidence is most useful for comparing models of the same data against one another. In isolation, $Z$ is the normalizing constant for $\Pi(\theta)$, ensuring the posterior is a well-defined probability distribution. $Z$ also represents the probability of generating the observed sample for all possible values of $\theta$, which we notice if we rearrange Bayes' theorem:

$$
Z = \int_{\forall \theta} L(\theta) \pi(\theta) d\theta
$$
Unlike other model selection methods that rely on the model's maximum likelihood estimate, evidence values are parameter-independent representations of a model's plausibility. This allows us to use $Z$ to compare models while accounting for differences in model complexity, such as through Bayes factors [@jeffreys1998]. Additionally, sharing evidence values for published models can be valuable for future researchers who may wish to build their own competing models.

### Introduction: Nested Sampling 

Technically, we could use a high-density posterior estimate generated by MCMC to calculate $Z$. This requires considerable extra computational beyond the non-trivial task of sampling from $\Pi(\theta)$; given this, evidence values are often ignored by researchers or treated as optional by-products, rather than as vitally important components of Bayesian inference [@skilling2006].

Nested sampling (NS), introduced in @skilling2004 and @skilling2006, is a Bayesian computational technique that instead focuses on estimating $Z$ directly while providing $\Pi(\theta)$ as an optional by-product. In addition, NS has several features that may make it more appealing than MCMC in certain scenarios, such as:

-   Clearly defined stopping criteria with no "burn-in" period required.

-   Ability to handle complex posteriors that typically confound MCMC approaches (e.g., multimodal distributions).

-   Uncertainty estimation from a single run without resampling.

NS approximates $Z$ by transforming the complex, multi-dimensional integral over $\pi(\theta)$ into a one-dimensional integral over nested "shells" of constant likelihood. These shells are ordered by likelihood values, with high-likelihood regions nested within low-likelihood regions. For a likelihood threshold $L^*$, the prior volume containing higher likelihoods is: $$V(L^*) = \int_{L(\theta) > L^*} \pi(\theta) d\theta$$

By sampling different $L^*$ values and their corresponding volumes $V(L^*)$, we can calculate the evidence as: $$Z = \int_0^1 L(V) dV $$ where $V = V(L^*)$ represents the prior volume associated with a given likelihood contour. If $L_1, ..., L_k$ represents our series of likelihoods and $V_1, ... V_k$ the corresponding series of volumes, we can estimate this integral with the trapezoidal method:

$$
Z \approx \sum_{i = 1}^{k} \frac{L_{i - 1} + L_i}{2} (V_{i - 1} - V_i)
$$

Since calculating $V$ directly is usually impossible, NS instead uses _likelihood-restricted prior sampling_ (LRPS) to ensure we can estimate the shrinkage in $V$ between subsequent iterations. The algorithm begins with $N$ randomly generated live points from the prior. At each iteration, the point with the lowest likelihood defines the current threshold $L^*$ and is removed. It is then replaced with a new point sampled from $\pi(\theta)$ subject to the constraint that its likelihood exceeds $L^*$. This process ensures that the prior volume shrinks predictably at each iteration $i$ [@speagle2020]:

$$
\begin{aligned}
  V_0 &= 1 \\
  V_i &= t_i V_{i-1} \\
  \Pr(t_i) &= Nt_i^{N - 1}
\end{aligned}
$$

where $t_i$ is the smallest value among $N$ draws from the standard uniform distribution.

NS provides a natural stopping criterion for a run: sampling terminates once the remaining unexplored prior volume contributes negligibly to our estimate of $Z$. After the run, we can use the contribution of each point to the evidence integral as weights to construct a sample from the posterior distribution.

## Nested Sampling with ernest

ernest attempts to make NS more familiar to statisticians most familiar with R and its system of packages. Unlike higher-level Bayesian inference packages, ernest makes the user responsible for specifying their model's likelihood function and prior distribution. To demonstrate how this is accomplished, here we will use ernest to find the evidence of the Poisson model we fit earlier using brms.

### Prior Distributions

Priors encode our knowledge of parameter distributions and are represented as `ernest_prior` objects. These can be created using `create_prior` or specialized functions like `create_normal_prior`.

Like most nested sampling implementations, ernest conducts its sampling within a $d$-dimensional unit cube and transforms points to the parameter space via a *prior transformation function*. This function should map uniform random variables from $[0,1]^d$ to the target prior distribution $\pi(\theta)$. For independent parameters, this transformation corresponds to the inverse cumulative distribution function (e.g., the quantile function in R `qnorm`).

Our model requires priors for regression coefficients; earlier, we specified that the intercept and linear parameters fall within a $N(0, 10)$ distribution. To provide this to ernest, we build our prior transformation function then wrap it with `create_prior`.
```{r}
ptf <- function(x) {
  # Transform each value in x from [0, 1] to N(0, 10)
  qnorm(x, mean = 0, sd = 10)
}

prior1 <- create_prior(ptf, names = c("Intercept", "zBase", "Trt1", "zBase_Trt1"))
prior1
```

We can use `create_prior` to build up more complicated priors as well. Consider if we wanted to truncate the intercept prior to be strictly non-negative: We could build two priors and then concatenate them together using `+`.

```{r}
ptf_int <- function(x) {
  # Transform the value in x from [0, 1] to N(0, 10) in [0, Inf]
  truncnorm::qtruncnorm(x, mean = 0, sd = 10, a = 0)
}

ptf_coef <- function(x) {
  # Transform each value in x from [0, 1] to N(0, 10)
  qnorm(x, mean = 0, sd = 10)
}

prior2_int <- create_prior(ptf_int, names = "Intercept")
prior2_coef <- create_prior(ptf_coef, names =  c("zBase", "Trt1", "zBase_Trt1"))
prior2 <- prior2_int + prior2_coef
```

In simpler cases involving independent normal or uniform priors, we can avoid writing a custom transformation function by using `create_normal_prior` or `create_uniform_prior`. We can take advantage of this for our Poisson model:

```{r}
fast_prior1 <- create_normal_prior(names = c("Intercept", "zBase", "Trt1", "zBase_Trt1"), sd = 10)
fast_prior1
```

### Likelihood Function

The log-likelihood function takes a vector of transformed model parameters and returns the corresponding log-likelihood value. For invalid parameter values, the function should return `-Inf`.

At present, it's out-of-scope for the ernest package to help users build and validate their likelihood function. For this, users should consult other resources or packages, such as [enrichwith](https://cran.r-project.org/web/packages/enrichwith/) and [tfprobability](https://cran.r-project.org/web/packages/tfprobability/index.html). For our Poisson model, we can use brms' extractor functions to help build an R function that calculates the log-likelihood: 

```{r}
data <- standata(fit1)
X <- data$X
Y <- data$Y
linkinv <- fit1$family$linkinv

ll_model <- function(theta) {
  predictors <- drop(X %*% theta)
  fitted_values <- linkinv(predictors)
  sum(dpois(Y, lambda = fitted_values, log = TRUE))
}
```

Optionally, we can use the `create_likelihood` function to specify how ernest should handle the likelihood function when it returns a non-finite, non-`-Inf` value. Should we ignore this step, ernest will warn us whenever such a value occurs.

### Likelihood-Restricted Prior Sampling

The central challenge in nested sampling is efficiently generating new points from the prior distribution that exceed a likelihood threshold. This ensures systematic exploration of progressively smaller, higher-likelihood regions of the parameter space. ernest provides several LRPS strategies for different problem types:

-   **Random Walk Metropolis-Hastings** (`rwmh_cube`): Performs random walks within the unit cube, accepting moves that satisfy the likelihood constraint. Originally proposed by @skilling2004, ernest's implementation includes automatic step-size tuning for user-specified acceptance rates.

-   **Slice Sampling within Hyperrectangles** (`slice_rectangle`): Draws from bounding hyperrectangles that shrink with each rejection. More efficient for complex or correlated parameter spaces.

-   **Uniform Sampling in Unit Cube** (`unif_cube`): Draws uniformly from the entire unit cube. Simple for testing components but generally inefficient for evidence estimation.

-   **Single Ellipsoidal Sampling** (`unif_ellipsoid`): Draws from the minimum-volume ellipsoid bounding the live points. Much more efficient than uniform sampling for unimodal distributions.

-   **Multi-Ellipsoidal Sampling** (`multi_ellipsoid`): Uses multiple ellipsoids to cover live points. Uniquely capable of detecting and adapting to multimodal posterior distributions.

### Setting Up and Running the Sampler

We can now perform our nested sampling run. First, we construct an `ernest_sampler` object: This object is bound to an environment that contains the current set of "live" points and their corresponding likelihood values.

```{r}
sampler_mod <- ernest_sampler(
  log_lik = ll_model,
  prior = fast_prior1,
  sampler = rwmh_cube(steps = 25),
  n_points = 300,
  seed = 42
)
sampler_mod
```

Nested samples are generated with `generate`. The `min_logz` argument sets the minimum log-ratio between the current (estimated) evidence and the remaining (estimated) evidence in the unexplored prior. Sampling ends once this ratio falls below `min_logz`, at which point its deemed that the remaining unexplored region of $\pi(\theta)$ contains a trivially small fraction of $Z$.

```{r}
run1 <- generate(sampler_mod, min_logz = 0.1)
run1
```

`run1` is an `ernest_run` object, containing the sampler used to generate the samples as well as its results. We can use this object to continue sampling from a given point, even after saving the original run to the disk.

```{r}
file_name <- tempfile("run1", fileext = c(".rds"))
saveRDS(run1, file_name)
rm(run1)

run1 <- readRDS(file_name)
run2 <- generate(run1, min_logz = 0.05)
run2
```

```{r include = FALSE}
unlink(file_name)
```

### Inspecting and Summarising Results

The result object has a `summary` method for viewing information about the run:

```{r}
summary(run2)
summary(run2)$run
```

The [posterior](https://CRAN.R-project.org/package=posterior) package offers methods for inspecting the points generated during a run. Calling `as_draws` on an `ernest_run` object produces a matrix with posterior log-weights calculated for each point. This matrix be re-sampled to produce a weighed estimate of the posterior distribution.

```{r}
weighted_post <- as_draws(run2) |>
  resample_draws()
summarise_draws(weighted_post, default_summary_measures())
```

To visualize the run's performance, we use the `plot` and `visualize` methods.

```{r}
plot(run2)
visualize(run2, type = "density")
visualize(run2, type = "trace")
```

## Comparing Evidence Values

*TODO.*

We can simulate the uncertainty of an NS run by generating random draws of the log-volume estimate at each iteration [@skilling2006, @higson2019]. This is done using `calculate`.

```{r}
calc_mod0 <- calculate(run2, ndraws = 500)
calc_mod0
```

## Final Notes

Nested sampling is a powerful technique that can help researchers choose parsimonious models, especially in challenging scenarios where traditional MCMC methods may struggle. However, nested sampling can be computationally intensive, and users should be aware of the following:

**Computational Complexity**: The computational cost of nested sampling scales with both the dimensionality of the parameter space and the desired precision of the evidence estimate. Each iteration requires drawing a new sample from the likelihood-constrained prior, which becomes increasingly difficult as the algorithm progresses and the likelihood constraint becomes more restrictive. In high-dimensional problems (typically \> 20-30 parameters), the efficiency of the chosen LRPS strategy becomes critical.

**Prior Spaces**: Evidence values will vary based on the scale of the prior distribution chosen for a nested sampling run. Choosing a wider prior distribution reduces the contribution of high-likelihood regions to the evidence estimate, resulting in a lower overall value. Therefore, we users should be careful in choosing priors that can be properly compared between models. In addition, users should be aware that the scale of the prior also directly impacts the overall expected runtime: The length of an NS run is proportional to the number of live points used (`n_points`) and the distance between the prior and posterior distributions.

The ernest package brings these advantages to R users with a flexible, extensible interface. It allows users to define custom priors and likelihoods, supports multiple LRPS strategies, and integrates seamlessly with popular R packages for model specification and posterior analysis. By using ernest, R users can perform robust Bayesian model comparison and inference with greater transparency and reproducibility, making it a valuable tool for applied Bayesian analysis.

------------------------------------------------------------------------

\emph{References}
